Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-11 ms/cover_letter/plosHIVcover_resubm2.tex
still working on ms/response

@@

@@

>>>>

\response{We have added some text here.}

@@

\response{
We agree that this causes some difficulties. Earlier drafts
actually focused on SPVL as the output measure, but we thought
that non-technical readers would have a better feeling for the
real-world significance of progression time; furthermore, because
progression time (unlike log(SPVL) is on a ratio scale, it becomes
reasonable to quote results in terms of proportional changes. 
We are taking the reviewer's advice and switching back to SPVL.
}

@@

\response{
We have changed the wording, and corrected a typo (the first value should have been 0.84 --- sorry!) As we say in our responses above, we have added more information about the choice of $r$, $I(0)$, $\alpha(0)$ in our sensitivity analyses (Figure 1).
}

### 2017-01-11 ms/ms.tex
still working on ms/response

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl\ using baseline parameters. Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution. Note that the initial condition has negligible effect on equilibrium virulence.
(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
(b) Effects of varying initial infectious density $I(0)$.
(c) Effects of varying initial mean virulence $\alpha(0)$.
The $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate.
The eco-evolutionary dynamics are qualitatively similar
across a range of initial conditions (panels b,c).}
\label{fig:panel3}
\end{figure}

@@

Choosing the initial conditions for the simulations is something of a
challenge. In the most common theoretical modeling contexts
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular quantitative choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, and an initial fraction infected of
$10^{-4}$, to match the parameters used by Shirreff \etal; we ran one
small set of sensitivity analyses with baseline values of the
epidemiological parameters (Table~1) to test the effects of the
initial conditions. Appendix TODO provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well. As $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl) 
and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).

>>>>

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl\ using baseline parameters. Each initial condition was varied independently in order to test their effects on transient dynamics of virulence evolution. Note that the initial condition has negligible effect on equilibrium virulence.
(a) Contrast between the three-stage Shirreff model and the single-stage model calibrated to varying initial exponential growth rates, $r$.
(b) Effects of varying initial infectious density $I(0)$.
(c) Effects of varying initial mean virulence $\alpha(0)$.
The middle $r=0.042$ (orange, dotted) curve in panel (a), calibrated to match the epidemic dynamics of Shirreff \etal's three-stage model \cite{shirreff_transmission_2011}, shows that our simplified single-stage HIV model can produce similar SPVL trajectories to the original three-stage model; the other $r$ values are chosen to show the effects of doubling or halving the initial epidemic growth rate. 
The eco-evolutionary dynamics are qualitatively similar
across a range of initial conditions (panels b,c).}
\label{fig:panel3}
\end{figure}

@@

Choosing the initial conditions for the simulations is something of a
challenge. In the most common theoretical modeling contexts
researchers are primarily interested in equilibria (or other long-term
dynamical attractors such as limit cycles) and are exploring models
that have a single stable attractor, so initial conditions can be
ignored as long as we run models for long enough.  In eco-evolutionary
dynamics, however, the initial conditions do affect our
conclusions. We have no empirical information that would justify a
particular quantitative choice of the fraction infected and the mean
and variance of the distribution of SPVL at the point when the
pandemic strain of HIV-1 entered the human population; in any case,
the level of realism of our model does not support such a detailed
consideration of the early dynamics of HIV.  In most cases, we started
with an initial \Lspvl\ of 3.0, to match the value used by
Shirreff \etal. Shirreff \etal\ use an initial $I(0)=10^{-3}$;
we show results for this value in \figurename~\ref{fig:panel3}.
Since some parameters in our sampling process (see next section)
lead to very low equilibrium prevalence, we switched to 
$I(0)=10^{-4}$ for the remainder of our analysis.
We ran one small set of sensitivity analyses with baseline values of the
epidemiological parameters (Table~1) to test the effects of the
initial conditions. Appendix~S1 provides
further details on initial conditions, such
as the initial distribution of SPVL around the mean and the
distribution of initial infected density across single people
and different partnership types.

@@

Our simplifications of Shirreff \etal's model \cite{shirreff_transmission_2011} reproduce its qualitative behaviour --- in particular, its predictions of virulence dynamics --- reasonably well. As we change $r$ from 0.42 to 0.84 (the former value matching the initial rate of increase in prevalence in Shirreff \etal's full model) the initial trajectory of increasing virulence brackets the rate from the original model (\figurename~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx 4.3$ vs. $\approx 4.6$ \Lspvl) 
and equilibrium virulence ($\approx 4.25$ vs. $\approx 4.5$ \Lspvl) than Shirreff's, even for matching initial incidence trajectories (i.e., $r=0.042~\textrm{year}^{-1}$).
